---
title: "Swordfern Mortality Sample Analyses"
output: html_document
date: "2025-11-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(kableExtra)
library(readr)
library(misty)
library(stringr)
library(BSDA)
library(car)
library(ggpubr)
library(statsExpressions)
library(RColorBrewer)
library(pillar)
library(readxl)
library(gridExtra)
library(rstatix)
library(Hotelling)
library(RVAideMemoire)
library(heplots)
library(effectsize)
```


```{r}
#setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Core/Server Files/swordfern")
data <- read.csv("~/GitHub/swordfern/data/lab/combined.results.csv")
```

```{r}
data <- rename(data,Tentative.Result = Tentative.Results..Internal.NCBI.Oomcycete.including.voucher.blast..x)
```


```{r}
data <- data %>% filter(common_name.x=="western sword fern")
```



```{r}
data$Tentative.Result[data$Tentative.Result=="Phytophthora europaea or abietivora?"] <- "Phytophthora sp."
data$Tentative.Result[data$Tentative.Result=="Phytophthora flexuosa?"] <- "Phytophthora sp."
```





```{r}
data$Project <- as.factor(data$Project)
data$place_guess.x <- as.factor(data$place_guess.x)
data$Oomycete.x <- as.factor(data$Oomycete.x)
```


```{r}
field.baits <- data %>% filter(Project=="Swordfern Soil Bait Isolates")
```

```{r}
roots <- data %>% filter(Project=="Swordfern Roots")
```


### Field Baits


## Summary Tables

## Number of Isolates


```{r}
ggplot(field.baits[field.baits$Tentative.Result!="" & !is.na(field.baits$Tentative.Result),],aes(Tentative.Result,fill=Genus.x)) +geom_histogram(stat="count")+coord_flip() +theme_bw() +guides(fill="none") +theme(axis.text = element_text(face = "italic")) + labs(y="Number of Isolates")
```

```{r}
levels(field.baits$Oomycete)
```


## Sample Collections

```{r}
sample.summary <- field.baits %>% summarize(n.sites = n_distinct(place_guess.x,na.rm=TRUE),n.samples=n_distinct(Sample),n.isolates=n(),n.oomycete.isolates=sum(Oomycete.x=="Oomycete",na.rm=TRUE),n.oomycete.spec=n_distinct(Tentative.Result),n.Phytoph.isolates=sum(Genus.x=="Phytophthora",na.rm=TRUE),n.Phytoph.spec=n_distinct(Tentative.Result[Genus.x=="Phytophthora" & !is.na(Tentative.Result)]),Phytoph.spp=paste(unique(Tentative.Result[Genus.x=="Phytophthora" & !is.na(Tentative.Result)]),collapse=", "))

kable(sample.summary,digits = 2) %>% kable_material(c("striped", "hover","condensed")) %>% kable_styling(full_width = F) %>% scroll_box(height = "400px")
```


## Phytophthora Isolates

```{r}
Phytophthora.summary <- field.baits %>% filter(Genus.x=="Phytophthora") %>% summarize(n.sites = n_distinct(place_guess.x,na.rm=TRUE),n.samples=n_distinct(Sample),n.isolates=n(),n.oomycete.isolates=sum(Oomycete.x=="Oomycete",na.rm=TRUE),n.oomycete.spec=n_distinct(Tentative.Result),n.Phytoph.isolates=sum(Genus.x=="Phytophthora",na.rm=TRUE),n.Phytoph.spec=n_distinct(Tentative.Result[Genus.x=="Phytophthora" & !is.na(Tentative.Result)]),Phytoph.spp=paste(unique(Tentative.Result[Genus.x=="Phytophthora" & !is.na(Tentative.Result)]),collapse=", "))

kable(Phytophthora.summary,digits = 2) %>% kable_material(c("striped", "hover","condensed")) %>% kable_styling(full_width = F) %>% scroll_box(height = "400px")
```



## Samples per 'place guess'


```{r}
site.summary <- field.baits %>% group_by(place_guess.x) %>% summarize(n.samples=n_distinct(Sample),n.isolates=n(),n.oomycete.isolates=sum(Oomycete.x=="Oomycete",na.rm=TRUE),n.oomycete.spec=n_distinct(Tentative.Result),n.Phytoph.isolates=sum(Genus.x=="Phytophthora",na.rm=TRUE),n.Phytoph.spec=n_distinct(Tentative.Result[Genus.x=="Phytophthora" & !is.na(Tentative.Result)]),Phytoph.spp=paste(unique(Tentative.Result[Genus.x=="Phytophthora" & !is.na(Tentative.Result)]),collapse=", "))
kable(site.summary,digits = 2) %>% kable_material(c("striped", "hover","condensed")) %>% kable_styling(full_width = F) %>% scroll_box(height = "400px")
```



```{r}
species.summary <- field.baits %>% filter(!is.na(Tentative.Result) & Tentative.Result!="") %>% group_by(Tentative.Result) %>% summarize(n.sites = n_distinct(place_guess.x,na.rm=TRUE),n.samples = n_distinct(Sample,na.rm=TRUE),n.isolates=n(),Genus.x=paste(unique(Genus.x))) 
species.summary
```




```{r}
ggplot(species.summary,aes(Tentative.Result,n.sites,fill=Genus.x)) +geom_col() +theme_bw() +coord_flip() + guides(fill=FALSE) +theme(axis.text = element_text(face = "italic")) +labs(x="Species",y="Number of Sites (e.g. Seward Park)")
```

```{r}
ggplot(species.summary,aes(Tentative.Result,n.samples,fill=Genus.x)) +geom_col() +theme_bw() +coord_flip() + guides(fill=FALSE) +theme(axis.text = element_text(face = "italic")) +labs(x="Species",y="Number of Samples")
```

```{r}
species.summary <- field.baits %>% filter(!is.na(Tentative.Result) & Tentative.Result!="") %>% group_by(site.location.description.x,Tentative.Result) %>% summarize(n.sites = n_distinct(place_guess.x,na.rm=TRUE),n.samples = n_distinct(Sample,na.rm=TRUE),n.isolates=n(),Genus.x=paste(unique(Genus.x))) 
species.summary
```

```{r}
ggplot(species.summary,aes(Tentative.Result,n.sites,fill=Genus.x)) +geom_col() +facet_wrap(~site.location.description.x) +theme_bw() +coord_flip() + guides(fill=FALSE) +theme(axis.text = element_text(face = "italic")) +labs(x="Species",y="Number of Sites")
```


```{r}
names(field.baits)
str(field.baits$percent.of.plants..of.same.species..affected.within.sight.....y)
```


```{r}
ggplot(field.baits,aes(Tentative.Result,percent.of.plants..of.same.species..affected.within.sight.....y ,fill=Genus.x)) +geom_boxplot()+coord_flip()+theme_bw()+guides(fill="none")
```

```{r}
ggplot(field.baits,aes(Tentative.Result,Dieback.percent.x,fill=Genus.x)) +geom_boxplot()+coord_flip()+theme_bw()+guides(fill="none")
```
# Greenhouse data

## Seward Soil Trials

##Data Visualization

Visualize data for Seward soil trials by comparing weights and number of fronds (total, healthy, and unhealthy) between treatment groups (control - potting mix, sterilized, wild).

```{r}
greenhouse.data <- read.csv("~/GitHub/swordfern/data/greenhouse/Final-Greenhouse-Data-Combined.csv")
greenhouse.data <- rename(greenhouse.data, "Weight" = "Weight..g.")
greenhouse.data$Total.Fronds <- greenhouse.data$Healthy.Fronds + greenhouse.data$Crispy.Fronds

seward.data <- greenhouse.data %>% filter(Experiment == "Seward Soil")

seward.roots.plot <- ggplot(seward.data, aes(x = Treatment, y = Weight, fill = Treatment)) + geom_violin() +geom_boxplot(width = 0.1, fill = "white") + labs(title = "Root Mass Beween Soil Treatments (Trial 1 and 2)", y = "Mass (g)", x = "Soil Treatment") + theme(plot.title =  element_text(hjust = 0.5))
seward.roots.plot + scale_fill_brewer(palette = "Pastel1")

seward.healthy <- ggplot(seward.data, aes(x = Treatment, y = Healthy.Fronds, fill = Treatment)) + geom_violin() + geom_boxplot(width = 0.1, fill = "white") + labs(title = "Number of Healthy Fronds between Soil Treatments (Trial 1 and 2)", y = "Mass (g)", x = "Number of Fronds") + theme(plot.title = element_text(hjust = 0.5))
seward.healthy + scale_fill_brewer(palette = "Pastel1")

seward.crispy <- ggplot(seward.data, aes(x = Treatment, y = Crispy.Fronds , fill = Treatment)) + geom_violin() + geom_boxplot(width = 0.1, fill = "white") + labs(title = "Number of Crispy Fronds between Soil Treatments (Trial 1 and 2)", y = "Mass (g)", x = "Number of Fronds") + theme(plot.title = element_text(hjust = 0.5))
seward.crispy + scale_fill_brewer(palette = "Pastel1")

seward.total <- ggplot(seward.data, aes(x = Treatment, y = Total.Fronds , fill = Treatment)) + geom_violin() + geom_boxplot(width = 0.1, fill = "white") + labs(title = "Number of Total Fronds between Soil Treatments (Trial 1 and 2)", y = "Mass (g)", x = "Number of Fronds") + theme(plot.title = element_text(hjust = 0.5))
seward.total + scale_fill_brewer(palette = "Pastel1")
```
## Seward Trial 1

## Seward Trial 2

```{r}
seward2 <- seward.data %>% filter(Trial.x == "Seward Trial Two")

leveneTest(Total.Fronds~as.factor(Treatment), data=seward2, na.rm = TRUE)
leveneTest(Weight~as.factor(Treatment), data = seward2, na.rm = TRUE)
```
There is not enough evidence to reject equality of variances between treatment groups. 

```{r}
qqPlot(seward2$Weight)
qqPlot(seward2$Total.Fronds)
hist(seward2$Total.Fronds)
hist(seward2$Weight)

seward2 %>% shapiro_test(Total.Fronds, Weight)
```
Dry swordfern root weight data not normally distributed according to visual inspection using qqPlot and histogram, and the Shapiro-Wilk test for normality.

```{r}
set.seed(123)
seward2.npmv <- nonpartest(Total.Fronds | Weight ~ Treatment, data = seward2, permtest = TRUE, plots = F, tests = c(1, 0, 0, 0), releffects = TRUE)
seward2.npmv
```


## Seward Combined Trials (1 and 2)

### Check equality of variances (all treatment groups)

```{r}
leveneTest(Healthy.Fronds~as.factor(Treatment), data=seward.data, na.rm = TRUE)
```
There is enough evidence to reject equality of variances between treatment groups since our p-critical value of 0.03578 was greater than our p-value of 0.05. This is likely due to the potting mix group being much smaller than the sterilized and wild soil groups. Due to this discrepancy, we will explore the data comparing only the sterilized and wild groups.

### Check equality of variances (Wild and Sterilized groups only)

```{r}
seward.nopot <- seward.data %>% filter(Treatment != "Control - Potting Mix")
leveneTest(Healthy.Fronds~as.factor(Treatment), data=seward.nopot, na.rm = TRUE)
leveneTest(Crispy.Fronds~as.factor(Treatment), data=seward.nopot, na.rm = TRUE)
```
There is not enough evidence to reject equality of variances between the number of healthy and crispy fronds between treatment groups.

### Check normality

```{r}
qqPlot(seward.nopot$Total.Fronds)
qqPlot(seward.nopot$Weight)
hist(seward.nopot$Weight)
hist(seward.nopot$Total.Fronds)


seward.log <- seward.nopot %>% mutate(log_total = log1p(Total.Fronds),
                                        log_weight = log1p(Weight))

seward.test <- seward.log %>% dplyr::select(log_weight, log_total)
mshapiro.test(seward.test)
seward.nopot %>% shapiro_test(Total.Fronds, Weight)
```
qqPlot and histogram for the number of total fronds are visually normally distributed. qqplot and histogram for the dried root weight are visually not normally distributed, therefore we cannot perform a MANOVA. Instead, we will perform a PERMANOVA (Permutational Multivariate Analysis of Variance) as this is a non-parametric statistical test that does not require data to be normally distributed. This is further confirmed by performing the Multivariate Shapiro-Wilk normality test on the natural logarithm of one plus our sample data which suggests that our distribution of data for the dried root weight and total number of fronds is not normal (p-value = 5.389e-13 < 0.05).

## Run Nonparametric Inference for Multivariate Data on Seward dataset

```{r warning=FALSE}
#install.packages("npmv")
library(npmv)

set.seed(123)
seward.npmv <- nonpartest(seward.nopot$Total.Fronds | seward.nopot$Weight ~ Treatment, data = seward.nopot, permtest = TRUE, plots = F, tests = c(1, 0, 0, 0), releffects = TRUE)
seward.npmv

seward.ss <- ssnonpartest(Total.Fronds | Weight ~ Treatment, data = seward.nopot, alpha = .05)
```
After running the ANOVA type test with F approximation for each subseet, the soil treatment has a statistically significant difference on the total number of fronds and dried root mass in our swordfern plants (p-value = 0.016 < 0.05).
The probability that a randomly chosen plant from the sterilized soil group exhibits a greater number of fronds than a randomly chosen plant from the trial is 0.57292. The probability that a randomly chosen plant from the sterilized soil group exhibits a greater root mass than a randomly chosen plant from the trial is 0.60694. 