---
title: "illumina-analyses"
output: html_document
date: "2026-09-25"
---

#Load Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyverse)
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
#BiocManager::install("ShortRead")
#BiocManager::install("dada2")
#BiocManager::install("DECIPHER")
#BiocManager::install("phyloseq")
#BiocManager::install("rBLAST")
library(tibble)
library(Biostrings)
library(DECIPHER)
library(dada2)
library(ShortRead)
library(phyloseq)
library(rBLAST)
library(ggplot2)
#devtools::install_github("gmteunisse/fantaxtic")
library(fantaxtic)
```

After running DADA2 pipeline through Kamiak (WSU's HPC) rds and csv files were saved to be imported into RStudio for further statistical analyses. 

#Import datasets 

```{r}
seqtab.nochim <- readRDS("seqtab_nochim.rds")
taxonomy.test <- read.csv("taxonomy_test.csv")
```

## Create sample information dataframe for phyloseq object

```{r, echo=FALSE}

samples.out <- rownames(seqtab.nochim)
location <- sapply(strsplit(samples.out, "-"), `[`, 1)
health <- sapply(strsplit(samples.out, "-"), `[`, 2)
fhl.num <- sapply(strsplit(samples.out, "[[:digit:]]-"), `[`, 2)
fhl.num <- sapply(strsplit(fhl.num, "_R1.fastq.gz"), `[`, 1)

sample.info <- data.frame(Location = location, Health = health, FHL = fhl.num)
rownames(sample.info) <- samples.out
sample.info <- sample.info %>% filter(str_detect(Health, "Dying|Dead|Healthy"))
```

Simplify sample info dataframe to recategorize health condition score to only contain "unhealthy" and "healthy" classifications by combining the "dying" and "dead" conditions.

```{r}
samples.2health <- sample.info %>% 
  mutate(Health = case_when(
    Health == "Dying" ~ "Unhealthy",
    Health == "Dead" ~ "Unhealthy",
    TRUE ~ "Healthy"
))
```

## Create phyloseq object
Create phyloseq object for 2 and 3 health classification group and add column to count how many reads each ASV got.

```{r}
ps <- phyloseq(
  otu_table(seqtab_matched, taxa_are_rows = FALSE),
  sample_data(sample.info),
  tax_table(taxa_matched)
)

dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps
theme_set(theme_bw())

otu.table <- as.data.frame(otu_table(ps))
taxa_matched$Count <- colSums(otu.table)

taxa_matched <- as.matrix(taxa_matched)

ps.count <- phyloseq(
  otu_table(seqtab_matched, taxa_are_rows = FALSE),
  sample_data(sample.info),
  tax_table(taxa_matched)
)

ps.2health <- phyloseq(
  otu_table(seqtab_matched, taxa_are_rows = FALSE),
  sample_data(samples.2health),
  tax_table(taxa_matched)
)
```

